PEACE: {\underline P}arallel {\underline E}nvironment for {\underline A}ssembly and {\underline C}lustering of Gene {\underline E}xpression

### 2010-02-14 papers/NAR_web_server_2009/paper.tex
Modificatons to abstract.

\begin{abstract}
  We present \peace, a stand alone tool for the high-throughput {\it
    ab initio} clustering of transcript fragment sequences produced by
  Next Generation or Sanger Sequencing technologies.  Installed and
  run through an easily downloaded GUI, \peace\/ can process large
  data sets and group the fragments by gene association, achieving a
  greater sensitivity (without sacrificing runtime) than the competing
  clustering tools and applicable to bases as small as 75 bases.
  Through the GUI the user can collect cluster statistics and examine
  specific clusters for more comprehensive study.  \peace\/, used as a
  ``pre-assembly'' tool, enables any assembly tool to work with
  individual small clusters of closely related sequences, resulting in
  significant increases in sensitive and robustness to sequencing
  error (e.g. with a 52\% improvement in sensitivity over the cluster
  tool, \wcd\/, and a 184\% improvement in sensitivity over a leading
  assembly tool, \capthree, given a a 5\% rate of sequencing errors), as
  well as a reduction in runtime (with a 40\% speedup, as compared to
  \wcd, when applied to a 361 Mb set of Arabidopsis transcriptome fragments).
\end{abstract}

@@

, access to a transcriptome sequence was primarily
through the use of Expressed Sequence Tags (ESTs), single-pass cDNA
sequences derived from transcribed mRNAs and sequenced by Sanger
Sequencing technology.  More recently, Next Generation Sequencing
(NGS) technology has begun to rapidly replace Sanger Sequencing,
allowing for more comprehensive coverage of the transcriptome.  For
example, ESTs now being added to the GenBank dbEST are increasingly
the product of NGS technologies such as 454 pyrosequencing, which
enables the sequencing of novel and rare transcripts at a considerably
higher rate of coverage than Sanger Sequencing
\cite{Cheung2006,Emrich2007}.  From a computational perspective, this
is a mixed blessing: while NGS provides immense quantities of new
information, it also provides immensely larger data sets -- and thus a
need for faster, more efficient analysis algorithms.

@@

Given a set of transcript fragments sampled from across the genome, a
necessary first step of their analysis is that of clustering: separating
the fragments according to the transcript from which they were
derived.  Frequently performed implicitly by assembly tools,
clustering the data as a ``pre-assembly'' step has a number of
advantages.  Most significantly: performing this step will
allow the application of the assembly tool to individual clusters --
saving significant amounts of time \cite{Hazelhurst08a}.

However clustering is a computationally challenging problem.  Even
with the smaller number of ESTs produced using Sanger Sequencing, the
runtime and memory requirements to cluster based on pair-wise sequence
alignments make such an approach infeasible in practice.  The much
larger data-set size produced by NGS technologies exacerbates this
problem.  To deal with this, \peace\/ combines our own version of the $d^2$
alignment-free sequence distance function \cite{Hide94} and the
concept of a {\it minimum spanning tree} \cite{Prim57} to quickly and
accurately find clusters of ESTs expressed from the same gene without
reference to a sequenced genome.  Compared against competing tools for
the same purpose
\cite{Burke99,Slater00,Huang99,Parkinson02,Kalyanaraman03,Malde03,
  Ptitsyn05,Hazelhurst08a,Picardi09}, \peace\/ proves to be both more
sensitive and more robust to sequencing error without sacrificing
runtime.

@@

The $d^2$ distance measure used to assign edge weights is an
alignment-free measurement of sequence distance that can be calculated
significantly faster than a Smith-Waterman alignment \cite{Hide94}.
$d^2$ works by comparing the frequency of words (strings of a fixed
length) appearing in a limited region of each string.  Fragments
overlapping by a sufficient length will share neighborhoods of enough
similarity to ensure a small distance even in the presence of a
moderate number of base errors.  In practice we employ our own
variation of $d^2$, the {\it two-pass $d^2$ algorithm}, which
heuristically searches for a neighborhood of maximum similarity and
then finding the $d^2$ score within that neighborhood (see
Supplementary Materials for details).

@@

\peace\/ has been test on both simulated and real data from NGS and
Sanger Sequencing technologies, comparing results against those
produced by the \wcd\/ clustering tool \cite{Hazelhurst08a} and the
\capthree\/ assembly tool \cite{Huang99} (the latter of which
implicitly calculates a cluster in the process of assembly).  For our
simulation tests we used the {\bf ESTSim} tool \cite{Hazelhurst03} to
generate simulated transcript of varying length fragments under
different models of error (Supplementary Materials, Section~C.1),
generating the fragments from the list of 100 zebra fish genes used in
the \wcd\/ testing \cite{Hazelhurst08a}.  Tool parameters were taken
to match, as closely as possible, those used in the \wcd\/ study (see
supplementary materials).  Our primary methods of quality assessment
were {\it sensitivity} (the fraction of fragment pairs from the same
gene that were correctly clustered together) and {\it Type 1 error}
(the fraction of genes that were divided between clusters)
\cite{Wang04,Hazelhurst08a}.  Specificity was not an issue in
simulated tests, as both \peace and \wcd were completely successful
in separating unrelated ESTs (save when dealing with recently
duplicated genes).  Other measurements are also discussed in the
supplementary materials.  In Figure~\ref{SeT1} (left), reflecting
tests done on simulated Sanger EST set, we see a
significant improvement of \peace\/ (blue) over \wcd\/ (green) and
\capthree\/ (black) in its sensitivity to sequencing errors, while in
Figure~\ref{SeT1} (right) we see a comparable improvement in Type 1
error.  In Figure~\ref{singletons} we look at the number of singleton
clusters (fragments not joined to any cluster), which should not occur
in our simulated sets.  We again see significant improvement in \peace.
The sequential runtime of \peace\/ is slightly improved over that of
\wcd\/ (see Supplementary Materials, Figure~S4).

@@

In applying the tools to real data, we started with a set of
approximately $190,000$ clean Sanger ESTs derived from the {\it
  Chlamydomonas reinhardtii} genome \cite{Liang2008}.  To compute
sensitivity we used the {\it gmap} tool \cite{Wu05} to map the
fragment set to the genome, taking this as our reference clustering.
We see slight improvements in \peace\/ over \wcd\/ in both sensitivity
and Type 1 error (both significantly outperforming \capthree).  Using
the mouse data set used in Hazelhurst {\it et al.}
\cite{Hazelhurst08a} \peace\/ again shows slight improvements in
sensitivity and Type 1 error rates, with an 18\% speedup for \peace
(see Supplementary Materials, Section~D, for more details).

A comparative test of \peace against \wcd on NGS short-read sequences
was not feasible due to the implementation of \wcd.  In theory that
tool should be able to handle short-read sequences with the
appropriate parameter settings.  However, in the distributed code those
parameters seemed to be hard-coded, thus rendering testing impossible.
\peace was run on both simulated and real NGS fragment sets.  For the
former we found \peace to do well, achieving a sensitivity of 0.75
when simulating a sequencing error rate of 2\% (with comparable {\it
  Jaccard Index} value, which reflects both sensitivity {\it and}
specificity \cite{Kalyanaraman03,Hazelhurst08a}).  Runtime was
approxemently 17.6 minutes to process the 7.9 Mb of data when run in
sequential mode on a 3.2 GHz Intel Xeon CPU.

@@

Here we have presented \peace, a stand alone tool for the
high-throughput clustering of transcript fragments capable of dealing
with sequences as short as 75 bases.  \peace\/, available at
\href{http://www.peace-tools.org}{www.peace-tools.org}, is open-source
and managed through a user friendly GUI that enables both local and
remote installation and execution in sequential or parallel mode.
Based on a novel algorithm for the clustering of the fragments by gene
association, \peace\/ shows significant improvement in sensitivity,
without sacrificing runtime, over the competing \wcd\/ tool
\cite{Hazelhurst08a}, and shows an order of magnitude in improvement
over the clustering preformed in the course of assembly by the
\capthree\/ tool \cite{Huang99}.

As a clustering tool based on sequence distance, \peace\/ faces
certain inherent limitations. For example, \peace\/ cannot handle
duplicate genes; like \wcd\/, it is unable to separate clusters
corresponding to genes with a greater than 88\% similarity.
Similarly, natural biological effects such as the trans-splicing of
transcripts, or effects from poorly cleaned transcript data (e.g. the
failure to remove sequencing adapters or post-transcriptional
poly(A)/(T) tails), and the presence of low-complexity repeats can
cause similar effects in these clustering tools.  The problems can be
handled through the application of the assembler, and the ability to
apply any assembler to small cluster (as opposed to the data set as a
whole) results in a significant reduction in overall assembly time.

>>>>

\begin{abstract}
  We present \peace, a stand alone tool for the high-throughput {\it
    ab initio} clustering of transcript fragment sequences produced by
  Next Generation or Sanger Sequencing technologies, freely available
  from \href{http://www.peace-tools.org}{www.peace-tools.org} .
  Installed and run through an easily downloaded GUI, \peace\/ can
  process large data sets of transcripts fragments of length 50 bases
  or greater, grouping the fragments by gene association and achieving
  a greater sensitivity (without sacrificing runtime) than the
  competing clustering tools.  Once clusered, the user can use the GUI
  to to collect statistics and single out specific clusters for more
  comprehensive study.  \peace\/, when used as a ``pre-assembly''
  tool, enables any assembly tool to work with individual small
  clusters of closely related sequences.  The result is a significant
  increase in sensitivity, as well as improved robustness to
  sequencing error as well as a reduction in runtime.  For a 5\%
  sequencing error rate, we see a 52\% improvement in sensitivity over
  the leading cluster tool, \wcd\/, a 184\% improvement in sensitivity
  over the assembly tool \capthree.  When applied to a 361 Mb set of
  Arabidopsis transcriptome fragments, we see a 40\% speedup over
  \wcd.
\end{abstract}

@@

In the past, access to a transcriptome sequence was primarily
through the use of Expressed Sequence Tags (ESTs), single-pass cDNA
sequences derived from transcribed mRNAs and sequenced by Sanger
Sequencing technology.  More recently, Next Generation Sequencing
(NGS) technology has begun to rapidly replace Sanger Sequencing,
allowing for more comprehensive coverage of the transcriptome.  For
example, ESTs now being added to the GenBank dbEST are increasingly
the product of NGS technologies such as 454 pyrosequencing, which
enables the sequencing of novel and rare transcripts at a considerably
higher rate of coverage than Sanger Sequencing
\cite{Cheung2006,Emrich2007}.  From a computational perspective, this
is a mixed blessing: while NGS provides immense quantities of new
information, it also provides immensely larger data sets -- and thus a
need for faster, more efficient analysis algorithms.

@@

Given a set of transcript fragments sampled from across the genome, a
necessary first step of the set's analysis is that of clustering: separating
the fragments according to the transcript from which they were
derived.  Frequently performed implicitly by assembly tools,
clustering the data as a ``pre-assembly'' step has a number of
advantages.  Most significantly: performing this step will
allow the application of the assembly tool to individual clusters --
saving significant amounts of time \cite{Hazelhurst08a}.

However, clustering is a computationally challenging problem.  Even
with the smaller number of ESTs produced using Sanger Sequencing, the
runtime and memory requirements to cluster on the basis of pair-wise
sequence alignments make such an approach infeasible in practice.  The
much larger data-set size produced by NGS technologies exacerbates
this problem.  To deal with this, \peace\/ combines our own version of
the $d^2$ alignment-free sequence distance function \cite{Hide94} and
the concept of a {\it minimum spanning tree} \cite{Prim57} to quickly
and accurately find clusters of ESTs expressed from the same gene
without reference to a sequenced genome.  Compared against \wcd, the
leading cluster tool in the literature \cite{Hazelhurst08a}, as well as
other tools designed for the same purpose 
\cite{Burke99,Slater00,Huang99,Parkinson02,Kalyanaraman03,Malde03,
  Ptitsyn05,Hazelhurst08a,Picardi09}, \peace\/ proves to be both more
sensitive and more robust to sequencing error without sacrificing
runtime.  Nor are of these tools designed for ease of installation or
use to the extent of \peace.

@@

The $d^2$ distance measure used to assign edge weights is an
alignment-free measurement of sequence distance that can be calculated
significantly faster than a Smith-Waterman alignment \cite{Hide94}.
$d^2$ works by comparing the frequency of words (strings of a fixed
length) appearing in a limited region of each string.  Fragments
overlapping by a sufficient length will share neighborhoods of enough
similarity to ensure a small distance even in the presence of a
moderate number of base errors.  In practice we employ our own
variation of $d^2$, the {\it two-pass $d^2$ algorithm}, which
heuristically searches for a neighborhood of maximum similarity and
then finds the $d^2$ score based on that neighborhood (see
Supplementary Materials for details).

@@

\peace\/ has been test on both simulated and real data from NGS and
Sanger Sequencing technologies, comparing results against those
produced by the \wcd\/ clustering tool \cite{Hazelhurst08a} and the
\capthree\/ assembly tool \cite{Huang99} (the latter of which
implicitly calculates a cluster in the process of assembly).  For our
simulation tests we used the {\bf ESTSim} tool \cite{Hazelhurst03} to
generate simulated transcript of varying length fragments under
different models of error (Supplementary Materials, Section~C.1),
generating the fragments from the list of 100 zebra fish genes used in
the \wcd\/ testing \cite{Hazelhurst08a}.  Tool parameters were taken
to match, as closely as possible, those used in the \wcd\/ study (see
supplementary materials).  Our primary methods of quality assessment
were {\it sensitivity} (the fraction of fragment pairs from the same
gene that were correctly clustered together) and {\it Type 1 error}
(the fraction of genes that were divided between clusters)
\cite{Wang04,Hazelhurst08a}.  Specificity was not an issue in
simulated tests, as both \peace\/ and \wcd\/ were completely successful
in separating unrelated ESTs (save when dealing with recently
duplicated genes).  Other measurements are also discussed in the
supplementary materials.  In Figure~\ref{SeT1} (left), reflecting
tests done on simulated Sanger EST set, we see a
significant improvement of \peace\/ (blue) over \wcd\/ (green) and
\capthree\/ (black) in its sensitivity to sequencing errors, while in
Figure~\ref{SeT1} (right) we see a comparable improvement in Type 1
error.  In Figure~\ref{singletons} we look at the number of singleton
clusters (fragments not joined to any cluster), which should not occur
in our simulated sets.  We again see significant improvement in \peace.
The sequential runtime of \peace\/ is slightly improved over that of
\wcd\/ (see Supplementary Materials, Figure~S4).

@@

In applying the tools to real data, we started with a set of
approximately $190,000$ clean Sanger ESTs derived from the {\it
  Chlamydomonas reinhardtii} genome \cite{Liang2008}.  To compute
sensitivity we used the {\it gmap} tool \cite{Wu05} to map the
fragment set to the genome, taking this as our reference clustering.
We see slight improvements in \peace\/ over \wcd\/ in both sensitivity
and Type 1 error (both significantly outperforming \capthree).  Using
the mouse data set used in Hazelhurst {\it et al.}
\cite{Hazelhurst08a} \peace\/ again shows slight improvements in
sensitivity and Type 1 error rates, with an 18\% speedup for \peace\/ 
(see Supplementary Materials, Section~D, for more details).

A comparative test of \peace\/ against \wcd\/ on NGS short-read sequences
was not feasible due to the implementation of \wcd.  In theory that
tool should be able to handle short-read sequences with the
appropriate parameter settings.  However, in the distributed versions those
parameters seemed to be hard-coded, thus rendering testing impossible.
\peace\/ was run on both simulated and real NGS fragment sets.  For the
former we found \peace\/ to do well, achieving a sensitivity of 0.75
when simulating a sequencing error rate of 2\% (with comparable {\it
  Jaccard Index} value, which reflects both sensitivity {\it and}
specificity \cite{Kalyanaraman03,Hazelhurst08a}).  Runtime was
approximately 17.6 minutes to process the 7.9 Mb of data when run in
sequential mode on a 3.2 GHz Intel Xeon CPU.

@@

Here we have presented \peace, a stand alone tool for the
high-throughput clustering of transcript fragments capable of dealing
with sequences as short as 50 bases.  \peace\/, available at
\href{http://www.peace-tools.org}{www.peace-tools.org}, is open-source
and managed through a user friendly GUI that enables both local and
remote installation and execution in sequential or parallel mode.
Based on a novel algorithm for the clustering of the fragments by gene
association, \peace\/ shows significant improvement in sensitivity,
without sacrificing runtime, over the competing \wcd\/ tool
\cite{Hazelhurst08a}, and shows an order of magnitude in improvement
over the clustering preformed in the course of assembly by the
\capthree\/ tool \cite{Huang99}.

As a clustering tool based on sequence distance, \peace\/ faces
certain inherent limitations. For example, \peace\/ cannot handle
duplicate genes; like \wcd\/, it is unable to separate clusters
corresponding to genes with a greater than 88\% similarity.
Similarly, other natural biological effects (e.g. the trans-splicing of
transcripts), effects from poorly cleaned transcript data (e.g. the
failure to remove sequencing adapters or post-transcriptional
poly(A)/(T) tails), and the presence of low-complexity repeats can
cause similar effects in these clustering tools.  The problems can be
handled through the application of the assembler, and the ability to
apply any assembler to small cluster (as opposed to the data set as a
whole) results in a significant reduction in overall assembly time.
